Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-03-31 ms.tex
fixed 6.2

Based on our numerical results and the recent work of \citet{McCourt+10}) we define thermal stability according to the criterion $\dot{q}_{\rm heat} > 10|\dot{q}_{\rm rad}|$ ($t_{\rm
  cool} > 10t_{\rm ff}$) being satisfied near the stagnation radius {\bf BDM: decide here}.  This condition translates into a
critical minimum heating rate
\begin{align}
v_{w} > v_{\rm TI} \simeq
  {\rm max}\begin{cases}
   200 \eta_{0.02}^{0.14} M_{\bullet,8}^{0.11}\,{\rm km\,s^{-1}}  &, \text{core}\\
   200 \eta_{0.02}^{0.17} M_{\bullet,8}^{0.08}\,{\rm km\,s^{-1}}   &, \text{cusp}  \\
\sigma_0(\zeta_c^{2}-1)^{1/2} \approx \sigma_0(r_{b}/r_{\rm soi})^{0.5(1-\Gamma)}, 
  \end{cases}
  \label{eq:cooling3}
\end{align}
where the final condition corresponds to the low wind heating limit (eq.~[\ref{eq:zetacrit}]), for which the stagnation radius moves out to the break radius $r_b$, resulting in thermal instability.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{mdot_sfr.pdf}
\caption{\label{fig:bh_growth} {\it Top panel:} Critical values of the
  black hole accretion rate $\dot{M}$ in Eddington units as a function
  of black hole mass $M_{\bullet}$, shown separately for core
  ($\Gamma=0.1$, solid lines) and cusp galaxies ($\Gamma=0.8$,
  dot-dashed lines).  Black lines show the accretion rate calculated
  from equation (\ref{eq:eddr_analytic}) using our estimate of the gas
  heating rate provided at $z = 0$ by the average star formation
  histories for each galaxy mass (Fig.~\ref{fig:vwSources}).  Green
  lines show the maximum accretion set by Ia supernovae blow-out
  (\ref{eq:eddr_Ia}).  Red lines show the accretion rate so obtained
  if Compton heating alone was acting (eq.~[\ref{eq:MdotC}]).  Yellow
  lines show the maximum accretion rate for a thermally stable flow
  near the stagnation radius (eq.~[\ref{eq:Mdotmax}]).  We also show
  Eddington ratios inferred for the galactic center from
  \citet{Quataert:2004a} (line with black circle) and for NGC3115 from
  \citet{ShcherbakovWong+:2014a} (line with black diamonds) and for
  the Swift J1644+57 event from \citet{BergerZauderer+:2012a} (line
  with black crosses) {\it Bottom panel:} Growth times $\Mbh/\Mdot$
  in units of the Hubble time $\th$ for each of the accretion rates
  shown in the top panel.}
\end{figure}

@@

Perhaps a more readily addressable test is whether a steady-state accretion
picture can account for the the large scatter, typically of $2-3$ orders of magnitude, in $L_{X}/L_{\rm
  edd}$ at fixed $M_{\bullet}$.  Scatter could result from at 
least three effects: (1) intrinsic scatter in the scaling relationships ($\Mbh-\sigma$ and $\Mbh-L_{\rm
  bulge}$) used to measure $\Mbh$; (2) the strong sensitivity of the accretion rate to
the stellar wind velocity, $\dot{M}/\dot{M}_{\rm edd} \propto
v_{w}^{-3.8(-2.4)}$ for core(cusp) galaxies, respectively
(Fig.~\ref{fig:mdot_mass}; eq.~[\ref{eq:eddr_analytic}]).  When 
combined with the significant dependence of $v_w$ on stochastic
intermittency in the star formation history (Appendix
\ref{fig:NickPlot2}), this can lead to order of magnitude differences
in $\dot{M}$; (3) differences angular momentum of the infalling gas
from galaxy to galaxy, resulting in differences in the fraction of the
gas lost to outflows, $\Mbh$.  Also potentially contributing is the
order of magnitude difference, at fixed $v_w$ and $M_{\bullet}$,
between $\dot{M}$ for core and cusp galaxies
(Fig.~\ref{fig:bh_growth}).  Differences in $\dot{M}$ are augmented by
the expectation that $L_{\rm X} \propto \dot{M}^{2} $ for radiatively
inefficient flows {\bf AG: Don't forget differences in eta}.

@@

Figure \ref{fig:bh_growth} (top panel) also shows that across all
black hole masses, Compton heating also cannot prevent average
accretion rates which are thermally unstable, i.e. $\dot{M}_{\rm C} >
\dot{M}_{\rm TI}$.  In the case of thermal instabilities, nuclear gas
may undergo cyclic oscillations of rapid cooling and high accretion
rates, followed by quiescent periods once gas has been consumed and/or
AGN feedback becomes effective.  Such cycles are proposed to occur on
large scales and long timescales in elliptical galaxies
(\citealt{Ciotti+10}), an idea possibly supported by, e.g., GALEX
observations showing that a small fraction of early-type galaxies in
the local universe have undergone recent ($< 1-2$ Gyr old) star
formation episodes (\citealt{Donas+07}).  Similar cyclic AGN activity
have been proposed to occur on the smaller scale of the sphere of
influence (e.g.~\citealt{Yuan&Li11}, \citealt{Cuarda+15}).

@@

\citet{BergerZauderer+:2012a} infer a flattening of the gas density
profile on radial scales $r \gtrsim 0.4$ pc (their Fig.~6) which looks
qualitatively similar to that the density profile we predict for core
galaxies (Fig.~\ref{fig:profiles}, solid line), although to obtain
such a flattening on the inferred radial scale, would require a large
$\Mbh\sim 10^7 \Msun$.  Their inferred gas profile also shows a break
at radii $\sim 0.6$ pc, which in our model corresponds to the break
radius of the galaxy and hence would imply an unrealistically small
SMBH mass $\Mbh\simeq 100\Msun$.  according to the mean relationship
from the \citet{Lauer+Faber:2007a} sample. A more likely explanation
is that the observed break in the gas density is associated with a
break in a nuclear star cluster in the galaxy {\bf AGs-ref Carson et
  al. 2015.}.

>>>>

Based on our numerical results and the recent work of \citet{McCourt+12}) we define thermal stability according to the criterion $\dot{q}_{\rm heat} > 10|\dot{q}_{\rm rad}|$ ($t_{\rm
  cool} > 10t_{\rm ff}$) being satisfied near the stagnation radius {\bf BDM: decide here}.  This condition translates into a
critical minimum heating rate
\begin{align}
v_{w} > v_{\rm TI} \simeq
  {\rm max}\begin{cases}
   200 \eta_{0.02}^{0.14} M_{\bullet,8}^{0.11}\,{\rm km\,s^{-1}}  &, \text{core}\\
   200 \eta_{0.02}^{0.17} M_{\bullet,8}^{0.08}\,{\rm km\,s^{-1}}   &, \text{cusp}  \\
\sigma_0(\zeta_c^{2}-1)^{1/2} \approx \sigma_0(r_{b}/r_{\rm soi})^{0.5(1-\Gamma)}, 
  \end{cases}
  \label{eq:cooling3}
\end{align}
where the final condition corresponds to the low wind heating limit (eq.~[\ref{eq:zetacrit}]), for which the stagnation radius moves out to the break radius $r_b$, resulting in thermal instability.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{mdot_sfr.pdf}
\caption{\label{fig:bh_growth} {\it Top panel:} Critical values of the
  black hole accretion rate $\dot{M}$ in Eddington units as a function
  of black hole mass $M_{\bullet}$, shown separately for core
  ($\Gamma=0.1$, solid lines) and cusp galaxies ($\Gamma=0.8$,
  dot-dashed lines).  Black lines show the accretion rate calculated
  from equation (\ref{eq:eddr_analytic}) using our estimate of the gas
  heating rate provided at $z = 0$ by the average star formation
  histories for each galaxy mass (Fig.~\ref{fig:vwSources}).  Green
  lines show the maximum accretion set by Ia supernovae blow-out
  (\ref{eq:eddr_Ia}).  Red lines show the accretion rate so obtained
  if Compton heating alone was acting (eq.~[\ref{eq:MdotC}]).  Yellow
  lines show the maximum accretion rate for a thermally stable flow
  near the stagnation radius (eq.~[\ref{eq:Mdotmax}]).  We also show
  Eddington ratios inferred for SgrA* from
  \citet{Quataert:2004a} (line with black circle), for NGC3115 from
  \citet{ShcherbakovWong+:2014a} (line with black diamonds), and for
  the Swift J1644+57 TDE from \citet{BergerZauderer+:2012a} (line
  with black crosses) {\it Bottom panel:} Growth times $\Mbh/\Mdot$
  in units of the Hubble time $\th$ for each of the accretion rates
  shown in the top panel.}
\end{figure}

@@

Perhaps a more readily addressable test is whether a steady-state accretion
picture developed in this work can account for the the large scatter, typically of $2-3$ orders of magnitude, in $L_{X}/L_{\rm
  edd}$ at fixed $M_{\bullet}$.  Scatter could result from at 
least three effects: (1) intrinsic scatter in the scaling relationships ($\Mbh-\sigma$ and $\Mbh-L_{\rm
  bulge}$) used to measure $\Mbh$; (2) the strong sensitivity of the accretion rate to
the stellar wind velocity and mass loss parameter, $\dot{M}/\dot{M}_{\rm edd} \propto
\eta v_{w}^{-3.8(-2.4)}$ for core(cusp) galaxies, respectively
(Fig.~\ref{fig:mdot_mass}; eq.~[\ref{eq:eddr_analytic}]).  When 
combined with the significant dependence of $v_w$ on stochastic
intermittency in the star formation history (Appendix
\ref{fig:NickPlot2}), this can lead to order of magnitude differences
in $\dot{M}$; (3) differences angular momentum of the infalling gas
from galaxy to galaxy, resulting in differences in the fraction of the
gas lost to outflows, $\Mbh$.  Also potentially contributing is the
order of magnitude difference, at fixed $v_w$ and $M_{\bullet}$,
between $\dot{M}$ for core and cusp galaxies
(Fig.~\ref{fig:bh_growth}).  Differences in $\dot{M}$ are augmented by
the expectation that $L_{\rm X} \propto \dot{M}^{2} $ for radiatively
inefficient flows.

@@

Figure \ref{fig:bh_growth} (top panel) also shows that across all
black hole masses, Compton heating also cannot prevent average
accretion rates which are thermally unstable, i.e. $\dot{M}_{\rm C} >
\dot{M}_{\rm TI}$.  In the case of thermal instabilities, nuclear gas
may undergo cyclic oscillations of rapid cooling and high accretion
rates, followed by quiescent periods once gas has been consumed and/or
AGN feedback becomes effective.  Such cycles are proposed to occur on
large scales and long timescales in elliptical galaxies
(\citealt{Ciotti+10}), an idea possibly supported by, e.g., GALEX
observations showing that a small fraction of early-type galaxies in
the local universe have undergone recent ($< 1-2$ Gyr old) star
formation episodes (\citealt{Donas+07}).  Similar cyclic AGN activity
have been proposed to occur on the smaller scale of the sphere of
influence (e.g.~\citealt{Yuan&Li11}, \citealt{Cuadra+15}).

@@

\citet{BergerZauderer+:2012a} infer a flattening of the gas density
profile on radial scales $r \gtrsim 0.4$ pc (their Fig.~6) which looks
qualitatively similar to that the density profile we predict for core
galaxies (Fig.~\ref{fig:profiles}, solid line), although to obtain
such a flattening on the inferred radial scale, would require a large
$\Mbh\sim 10^7 \Msun$.  Their inferred gas profile also shows a break
at radii $\sim 0.6$ pc, which in our model corresponds to the break
radius of the galaxy and hence would imply an unrealistically small
SMBH mass $\Mbh\simeq 100\Msun$.  according to the mean relationship
from the \citet{LauerFaber+:2007a} sample. A more likely explanation
is that the observed break in the gas density is associated with a
break in a nuclear star cluster in the galaxy {\bf AGs-ref Carson et
  al. 2015.}.
